---
title: "HW_5(Markov)"
author: "Emery Berry"
format: html
editor: visual
---

## Question 2

#### Part b) Find stationary distribution numerically

```{r}
a <- 0.04; b <- 0.16; K <- 0.1
states <- 1:5
n <- 5

# Build transition matrix
P <- matrix(0, n, n)
for (i in 1:n) {
  p_fwd <- K * exp(a * i)
  q_bwd <- K * exp(b * (i - 1))
  if (i < n) P[i, i+1] <- p_fwd
  if (i > 1) P[i, i-1] <- q_bwd
  P[i, i] <- 1 - sum(P[i,])
}

eig <- eigen(t(P))
idx <- which.min(abs(eig$values - 1))
pi_b <- Re(eig$vectors[, idx])
pi_b <- pi_b / sum(pi_b)
print(pi_b)
```

#### Part c) Simulate Markov Chain

```{r}
set.seed(42)
n_steps <- 1e6
chain <- integer(n_steps + 1)
chain[1] <- 1

for (t in 1:n_steps) {
  chain[t+1] <- sample(states, 1, prob = P[chain[t], ])
}

pi_sim <- tabulate(chain[-1], nbins = n) / n_steps

# Plot histogram of fraction of time spent in each state
barplot(pi_sim,
        names.arg = states,
        col = "pink",
        xlab = "State",
        ylab = "Fraction of time",
        main = "Fraction of time spent in each state (10^6 steps)")
```

#### Part d) Plot Histogram

```{r}
# Analytical from part (a)
pi_a <- exp((a - b) * states * (states - 1) / 2)
pi_a <- pi_a / sum(pi_a)

# Plot
barplot(
  rbind(pi_sim, pi_a, pi_b),
  beside = TRUE,
  names.arg = states,
  col = c("pink", "skyblue", "darkgreen"),
  xlab = "State", ylab = "Probability",
  main = "Stationary Distribution — Three Methods",
  legend.text = c("(c) Simulation", "(a) Analytical", "(b) Eigenvector"),
  args.legend = list(bty = "n")
)
```

-   Histograms Look to be overall matching!

# Question 3

#### Part b) Eigenvectors and Eigenvalues

```{r}
make_P <- function(a) {
  matrix(c(1-a,  a,   0,
             a,  0, 1-a,
             0, 1-a,  a),
         nrow = 3, byrow = TRUE)
}

# Leave a as a symbol — show the structure for a general a
# then compute numerically for a specific value
a_val <- 0.5  # or any valid 0 < a < 1
A <- t(make_P(a_val))

eig <- eigen(A)
cat("Eigenvalues:\n");  print(round(Re(eig$values), 6))
cat("Eigenvectors:\n"); print(round(Re(eig$vectors), 6))
```

#### Part c)

```{r}
a_val <- 0.99
P <- make_P(a_val)
A <- t(P)

# Get eigenvectors and eigenvalues
eig <- eigen(A)
lambdas <- Re(eig$values)
V <- Re(eig$vectors)  

cat("Eigenvalues:\n"); print(round(lambdas, 8))
cat("\nEigenvectors (columns):\n"); print(round(V, 6))

# X_0 = 1 means we start certainly in state 1
# so q_0 = (1, 0, 0) — all probability mass on state 1
q0 <- c(1, 0, 0)

# Find coefficients c such that V %*% c = q0
# i.e. express q0 in the eigenvector basis
coeffs <- solve(V, q0)
cat("\nCoefficients (c_1, c_2, c_3):\n"); print(round(coeffs, 8))

# q_n = c_1*lambda_1^n*v_1 + c_2*lambda_2^n*v_2 + c_3*lambda_3^n*v_3
# The lambda=1 term is the stationary distribution pi.
q_n <- function(n) {
  result <- rep(0, 3)
  for (k in 1:3) {
    result <- result + coeffs[k] * lambdas[k]^n * V[, k]
  }
  result
}

# Show q_n(1) converging to pi(1) = 1/3
cat("\nq_n(1) converging to pi(1) = 1/3:\n")
for (n in c(0, 1, 2, 5, 10, 25, 50, 100)) {
  cat(sprintf("  n = %3d:  q_n(1) = %.8f  |error| = %.2e\n",
              n, q_n(n)[1], abs(q_n(n)[1] - 1/3)))
}
```

Since $|\lambda_2|, |\lambda_3| < 1$ , the terms $c_2\lambda_2^n v_2$ and $c_3\lambda_3^n v_3$ decay to zero exponentially as $n \to \infty$, leaving only $c_1\lambda_1^n v_1 = c_1 v_1 = \pi$. Thus $q_n$ converges exponentially to the stationary distribution $\pi$, with convergence rate governed by $|\lambda_2|$: $q_n = c_1\lambda_1^n v_1 + c_2\lambda_2^n v_2 + c_3\lambda_3^n v_3 \xrightarrow{n \to \infty} \pi$

#### Part d)

```{r}
# Compute |lambda_2| across all valid values of a
a_seq <- seq(0.01, 0.99, length.out = 500)

lambda2 <- sapply(a_seq, function(a) {
  eigs <- abs(Re(eigen(t(make_P(a)))$values))
  sort(eigs, decreasing = TRUE)[2]  # second largest
})

opt_a <- a_seq[which.min(lambda2)]
cat(sprintf("Fastest convergence at a = %.4f\n", opt_a))
cat(sprintf("Minimum |lambda_2|      = %.6f\n", min(lambda2)))
```

#### Part e)

```{r}
eig <- eigen(A)
lambdas <- Re(eig$values)
V <- Re(eig$vectors)
coeffs <- solve(V, c(1, 0, 0))

q_n <- function(n) {
  result <- rep(0, 3)
  for (k in 1:3) {
    result <- result + coeffs[k] * lambdas[k]^n * V[, k]
  }
  result
}

# Theoretical q_n(1) for n = 0 to 300
n_max <- 300
theory <- sapply(0:n_max, function(n) q_n(n)[1])

# Simulate N chains, all starting at state 1
run_sim <- function(N_chains) {
  chains <- matrix(1L, nrow = N_chains, ncol = n_max + 1)
  for (t in 1:n_max) {
    chains[, t+1] <- sapply(chains[, t], function(s) sample(1:3, 1, prob = P[s, ]))
  }
  # fraction of chains in state 1 at each time n
  colMeans(chains == 1)
}

# Plot for N = 100, 1000, 10000
par(mfrow = c(1, 3))
for (N in c(100, 1000, 10000)) {
  set.seed(42)
  fn <- run_sim(N)
  plot(0:n_max, fn, type = "l", col = "skyblue", lwd = 1,
       xlab = "n", ylab = expression(f[n]),
       main = sprintf("N = %d chains", N),
       ylim = c(0, 1))
  lines(0:n_max, theory, col = "pink", lwd = 2, lty = 2)
  legend("topright",
         legend = c(expression(f[n]*" (simulation)"), expression(q[n](1)*" (theory)")),
         col = c("skyblue", "pink"),
         lty = c(1, 2), bty = "n")
}
par(mfrow = c(1, 1))
```
